---
title: Optimal Binning (Quantization)
author: Gaurav Sood
output: md_document
---

Quantization is an optimization problem \~ Minimize MSE/MAE given the number of bins by finding the optimal bin edges. Options:

1.  Solve the optimization problem

2.  Solve it via CART (which is \~ solving the said problem)

3.  Use percentile binning (\~ probability density)

4.  Use clustering methods like K-Means

5.  If the maximum number of bins was not fixed, we could use popular heuristic solutions for inferring k, e.g., Freedman-Diaconis and Sturges, and then we could use an optimization algorithm to find the optimal bin edges. Or we could use DBScan, etc.

Let's generate some normally distributed data.

```{r}
# Set the seed for reproducibility
set.seed(123)

# Simulate data from a unit normal distribution
data <- rnorm(10000)
```

Solve the optimization problem. 

```{r}
## Objective Function
# Define the objective function to minimize the MSE
mse_objective <- function(bin_edges) {
  bins <- cut(data, breaks = bin_edges, include.lowest = TRUE)
  binned_means <- tapply(data, bins, mean)
  mse <- mean((data - binned_means[bins])^2)
  return(mse)
}

# Set the number of bins
k <- 15

# Set the lower and upper bounds for the bin edges
lower_bound <- min(data)
upper_bound <- max(data)

# Add a small amount of noise to ensure unique bin edges
epsilon <- 1e-10
bin_edges <- seq(lower_bound, upper_bound, length.out = k+1) + runif(k+1, -epsilon, epsilon)

# Define the optimization problem
optim_result <- optim(par = bin_edges[-1], fn = mse_objective, method = "SANN")

# Get the optimal bin edges
optimal_bin_edges <- c(lower_bound, sort(optim_result$par), upper_bound)

# Calculate the binned data based on the mean of each bin
bins <- cut(data, breaks = optimal_bin_edges, include.lowest = TRUE)

# Calculate the bin means --- it can have NAs as some bins may have 0 entries
bin_means <- tapply(data, bins, mean)

# Map data points to the corresponding bin means
binned_data <- bin_means[as.numeric(bins)]

# Calculate the mean squared error (MSE)
mse <- mean((data - binned_data)^2)
cat("Mean Squared Error (MSE) for Optimal with K of 15:", mse, "\n")
```

2. CART solution [here](tree_split.ipynb) as R flakes. Bug in rpart.

3. Percentile binning

```{r}
# Set the number of percentiles (bins)
k <- 15

# Split the data into k percentiles
percentile_bins <- cut(data, breaks = quantile(data, probs = seq(0, 1, length.out = k + 1)), include.lowest = TRUE, labels = FALSE)

# Calculate the bin values as the mean within each percentile
bin_values <- tapply(data, percentile_bins, mean)

# Assign each data point to the corresponding bin value
binned_data <- bin_values[percentile_bins]

# Calculate the mean squared error (MSE)
mse <- mean((data - binned_data)^2)

# Print the MSE
cat("Mean Squared Error (MSE) for percentile =", mse, "\n")
```

4. K-Means

```{r}
# Apply k-means clustering
k <- 15  # Number of clusters
kmeans_result <- kmeans(data, centers = k)

# Get the cluster means
cluster_means <- kmeans_result$centers

# Assign each data point to the nearest cluster mean
nearest_cluster <- kmeans_result$cluster

# Assign to bin mean
binned_data <- cluster_means[nearest_cluster]

# Calculate the Mean Squared Error (MSE)
mse <- mean((data - binned_data)^2)

# Print the MSE
cat("Mean Squared Error (MSE) for kmeans for k = 15 =", mse)
```

5.  Using Freedman Diaconis Rule but with linear binning (which can result in bins w/ 0 entries) yield excellent results (though note the number of bins).

```{r}
library(MASS)

# Calculate the number of bins using the Freedman-Diaconis rule
num_bins <- nclass.FD(data)
cat("Number of bins:", num_bins, "\n")

# Calculate the bin edges
bin_edges <- seq(min(data), max(data), length.out = num_bins + 1)

# Calculate the binned data based on the mean of each bin
bins <- cut(data, breaks = bin_edges, include.lowest = TRUE)

# Calculate the bin means --- it can have NAs as some bins may have 0 entries
bin_means <- tapply(data, bins, mean)

# Map data points to the corresponding bin means
binned_data <- bin_means[as.numeric(bins)]

# Calculate the mean squared error (MSE)
mse <- mean((data - binned_data)^2)
cat("Mean Squared Error (MSE) for Freedman-Diaconis binning:", mse, "\n")
```

Let's find the optimal number of bins using Sturges' method.

```{r}
# Calculate the number of bins using Sturges' formula
num_bins <- ceiling(log2(length(data)) + 1)

cat("Number of bins:", num_bins, "\n")

# Calculate the bin edges using the range of the data
bin_edges <- seq(min(data), max(data), length.out = num_bins + 1)

# Slightly diff. way ...
bin_means <- sapply(1:(length(bin_edges) - 1), function(i) {
  mean(data[data >= bin_edges[i] & data < bin_edges[i + 1]])
})

binned_data <- sapply(data, function(value) {
  bin_index <- findInterval(value, bin_edges, rightmost.closed = TRUE)
  bin_means[bin_index]
})

mse_sturges <- mean((data - binned_data)^2)
print(paste("Mean Squared Error (MSE) for Sturges' formula:", mse_sturges))
```

DBScan.

DBScan is a disaster. Not only is it in that hyperparams like eps etc. need to be carefully chosen, we have to deal with noise points and find their bins.

```{r}
### DBScan
library(dbscan)  # Load the dbscan package

# Convert data to a matrix format
data_matrix <- matrix(data, ncol = 1)

# Set the DBSCAN parameters
eps <- 0.2  # Maximum distance between points in a cluster
minPts <- 5  # Minimum number of points to form a cluster

# Apply the DBSCAN algorithm
dbscan_result <- dbscan(data_matrix, eps = eps, minPts = minPts)

# Get the cluster assignments and noise points
cluster_assignments <- dbscan_result$cluster
noise_points <- dbscan_result$cluster == 0

# Calculate the cluster means (excluding noise points)
cluster_means <- tapply(data[!noise_points], cluster_assignments[!noise_points], mean)

# Calculate the distance between each data point and its corresponding cluster mean
distances <- mapply(function(d, b) (d - b)^2, data[!noise_points], cluster_means[cluster_assignments[!noise_points]])

# Convert noise data to a matrix format
noise_data_matrix <- matrix(data[noise_points], ncol = 1)

# Apply the DBSCAN algorithm on noise data
noise_to_cluster <- dbscan::dbscan(noise_data_matrix, eps = eps, minPts = 1)$cluster
noise_to_cluster_means <- tapply(data[noise_points], noise_to_cluster, mean)

# Calculate the distance between each noise point and its assigned cluster mean
noise_distances <- mapply(function(d, b) (d - b)^2, data[noise_points], noise_to_cluster_means[noise_to_cluster])

# Calculate the mean squared error (MSE)
mse <- mean(c(distances, noise_distances))
cat("Mean Squared Error (MSE) with DBScan:", mse, "\n")
```
